---
title: "HOMEWORK: Run Fuzzy design RDD"
title-block-banner: true
subtitle: "Mineure action publique"
author:
  - name: Olivier CARON
    email: caron.olivier.80@gmail.com
    affiliations: 
      name: "Paris Dauphine-PSL"
      city: Paris
      state: France
date: "last-modified"
toc: true
toc-depth: 3
toc-expand: 3
number-sections: true
number-depth: 10
format:
  html:
    self-contained: true
    theme:
      light: litera
    code-fold: true
    code-summary: "Display code"
    code-tools: true #enables to display/hide all blocks of code
    code-copy: true #enables to copy code
    grid:
      body-width: 800px
      margin-width: 100px
    toc: true
    toc-title: "Table of content"
    toc-location: left
execute:
  echo: true
  warning: false
  message: false
editor: visual
fig-align: "center"
highlight-style: ayu-light
reference-location: margin
#lightbox: true
filters:
  - parse-latex
---

## Load data {.unnumbered}

```{r}
library(arrow)
library(dplyr)
library(AER)      # For the ivreg function for 2SLS estimation
library(huxtable) # For table formatting
library(skimr)


# Load data
data <- arrow::read_parquet("data/EduloanFEMALE.parquet")

skim(data)
```

Below is one way to set up and run a fuzzy RDD using a 2SLS approach in R.

In this example we assume that the research question is to estimate the effect of `actually obtaining a loan` (the endogenous regressor) on `enrolling in a public university` (the outcome), where the probability of getting a loan is influenced by whether the applicant is above or below the cutoff (the instrument). In our dataset:

-   Outcome: **univ_register** (Enrolment in public university)

-   Endogenous Variable (Treatment): **loan** (Loan obtained)

-   Instrument: **discont_r** (Indicator for being above the cutoff)

-   Running variable control: **empirica_r** (the running variable re-centered at the threshold)

## Run the fuzzy RDD using 2SLS

```{r}
#| label: fuzzy_rdd
#Run the fuzzy RDD using 2SLS via ivreg
# Model specification:
#   Outcome: univ_register (university enrollment)
#   Endogenous regressor: loan (loan obtained)
#   Instrument: discont_r (discontinuity indicator)
#   Control: empirica_r (centered running variable)
fuzzy_rdd <- ivreg(univ_register ~ loan + empirica_r | discont_r + empirica_r, data = data)

# Display the summary of the 2SLS estimation
#summary(fuzzy_rdd)
#stargazer::stargazer(fuzzy_rdd, type="text")

hux_fuzzy <- huxreg(
  "Fuzzy RDD (2SLS)" = fuzzy_rdd,
  statistics = c("N" = "nobs", "R²" = "r.squared"),
  stars = c(`***` = 0.01, `**` = 0.05, `*` = 0.1)
)




```

## First-Stage Regression for Instrument Relevance

```{r}
#| label: endogeneity_test
# First-stage regression: the effect of the discontinuity on the loan variable
first_stage <- lm(loan ~ discont_r + empirica_r, data = data)
#summary(first_stage)
#stargazer::stargazer(first_stage, type="text")

huxtable::huxreg(
  "First Stage" = first_stage,
  statistics = c("N" = "nobs", "R²" = "r.squared"),
  stars = c(`***` = 0.01, `**` = 0.05, `*` = 0.1)
)

```

## Table of RDD Estimates

```{r}
#| label: huxtable
# Create huxtable object for publication-ready table
hux_fuzzy <- huxreg(
  "Fuzzy RDD (2SLS)" = fuzzy_rdd,
  "First Stage" = first_stage,
  statistics = c("N" = "nobs", "R²" = "r.squared"),
  stars = c(`***` = 0.01, `**` = 0.05, `*` = 0.1)
)

# more personalized table
hux_fuzzy <- hux_fuzzy %>%
  set_caption("Table 1: Fuzzy Regression Discontinuity Design (RDD) using 2SLS") %>%
  set_number_format(2) %>%
  set_font_size(11) %>%
  add_footnote("Notes: This table presents the results of a Fuzzy Regression Discontinuity Design (RDD) estimated using Two-Stage Least Squares (2SLS). The dependent variable is university enrollment. The endogenous variable is 'loan obtained', instrumented by the discontinuity indicator. Control variable: 'empirica_r'. Standard errors in parentheses. Significance levels: *** p<0.01, ** p<0.05, * p<0.10.")

# transform to latex code so I can ask ChatGPT to review it and use it below
latex_code <- to_latex(hux_fuzzy)

hux_fuzzy

```

## Test Latex

\documentclass{article}
\usepackage{booktabs}
\usepackage{threeparttable}

\begin{document}

\begin{table}[ht]
    \centering
    \caption{Fuzzy Regression Discontinuity Design (RDD) using 2SLS}
    \label{tab:fuzzy_rdd}
    \begin{threeparttable}
    \begin{tabular}{lcc}
        \toprule
        & \textbf{Fuzzy RDD (2SLS)} & \textbf{First Stage} \\
        \midrule
        (Intercept)   & 0.54 *** (0.02)  & 0.06 *** (0.01) \\
        Loan        & 0.21 *** (0.04)  & -- \\
        Empirica\_r & 0.00 (0.00)    & 0.00 *** (0.00) \\
        Discont\_r  & --                 & 0.45 *** (0.02) \\
        \midrule
        Observations & 5123 & 5123 \\
        R$^2$ & 0.05 & 0.24 \\
        \bottomrule
    \end{tabular}
    \begin{tablenotes}
        \item \textit{Notes:} This table presents the results of a Fuzzy Regression Discontinuity Design (RDD) estimated using Two-Stage Least Squares (2SLS). The dependent variable is university enrollment. The endogenous variable is `loan obtained`, instrumented by the discontinuity indicator. Control variable: `empirica_r`. Standard errors in parentheses. Significance levels: *** p$<$0.01, ** p$<$0.05, * p$<$0.10.
    \end{tablenotes}
    \end{threeparttable}
\end{table}

\end{document}
